### 2016-03-01 doc/paper/paper.tex
Added figures and started poster

\documentclass[12pt]{report}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear]{biblatex}
\usepackage{nicefrac}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage[acronym,nonumberlist]{glossaries}

\newcommand{\comment}[1]{\marginnote{\textit{#1}}}

@@

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}

@@

\frenchspacing
\onehalfspacing

\begin{document}

@@

From a theoretical perspective, network structure can profoundly influence the
incidence and prevalence trajectories of an epidemic, in turn affecting the
estimates of quantities such as effective population size. More practically,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{little2014using}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[but not all,
see][]{jones2003assessment}. This result has been coroborated by more recent
investigations using phylogenetic methods~\autocite{brown2011transmission}.
Under these models, nodes with a high number of contacts attract new
connections at an elevated rate. Networks produced by preferential attachment
have a power-law degree distribution, meaning that the number of nodes of
degree $k$ is proportional to $k^\gamma$ for some constant $\gamma$. When $2 <
\gamma \leq 3$, the network is referred to as ``scale-free''. The first contact
network model incorporating preferential attachment was introduced by
\textcite{barabasi1999emergence}, and is now referred to as the \gls{BA} model.
Under this model, networks are formed by iteratively adding nodes with $m$ new
edges each. These new edges are joined to existing nodes of degree $k$ with
probability proportional to $k^\alpha$, so that nodes of high degree tend to
attract more connections (in the original paper, only $\alpha = 1$ was
investigated).

Due to their complexity, it is generally difficult to explicitly calculate the
likelihood of an observed transmission tree under a contact network model. To
calculate the normalizing constant, we would need to integrate over all
possible networks, and also over all possible labellings of the internal nodes
of the transmission tree. While it is not known (to us) whether such
integration is tractable, a simpler alternative is offered by likelihood-free
methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
calculating the likelihood may be impossible, generating simulated datasets
according to a model is often straightforward. If our model fits the data well,
the simulated data it produces should be similar to the observed data. More
formally, if $D$ is the observed data, the posterior distribution $f(\theta
\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
specific case when $\rho$ is a kernel function, the approach is known as
kernel-\gls{ABC}.

Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
contact network parameters from an estimated transmission tree. We then
estimate the parameters of the \gls{BA} model on a variety of simulated and
real data sets.

@@

We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics and transmission trees over static contact networks,
as has been done previously \autocite[\textit{e.g.}][]{o2010contact,
robinson2013dynamics, leventhal2012inferring, groendyke2011bayesian}. To check
that our implementation was correct, we reproduced Figure 1 of
\textcite{leventhal2012inferring} (see Figure~\ref{fig:sf1}).

We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
denotes the total number of nodes in the network, and $I$, which is the number
of infected nodes at which to stop the simulation and sample the transmission
tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
they became infected at a rate proportional to their numbers of infected
neighbours, and never recovered. For all analyses, the transmission trees'
branch lengths were scaled by dividing by their mean. We used the 
\software{igraph} library's implementation of the \gls{BA}
model~\autocite{csardi2006igraph} to generate the graphs.

@@

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. To ensure our implementation was correct, we
applied it to the same mixture of Gaussians used by
\citeauthor{del2012adaptive} to demonstrate their method (originally used
by~\textcite{sisson2007sequential}). We were able to obtain a close
approximation to the function (see Figure~\ref{fig:smctest}), and attained the
stopping condition used by the authors in a comparable number of steps.

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0, 0.5, 1, 1.5,
2}. The adaptive \gls{ABC} algorithm was applied to each tree with these
priors: $N$ = Uniform(500, 15000), $I$ = Uniform(500, 5000), $m$ = Uniform(1,
5), and $\alpha$ = Uniform(0, 2). Following \textcite{del2012adaptive} and
\textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
equal to twice the empirical variance of the particles. Because there must be
at least as many nodes in the network as infected nodes, the proposal ratio was
set to zero whenever $I > N$. The algorithm was run with 1000 particles, 5
simulated datasets per particle, and the ``quality'' parameter set to 0.95. We
use the same stopping criterion as \textcite{del2012adaptive}, namely when the
MCMC acceptance rate dropped below 1.5\%. Point estimates for the parameters
were obtained by taking the highest point of an estimated kernel density on the
final set of particles.

@@

We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
model generates networks with a single connected component, we specifically
searched for datasets which originated from existing clusters, either
phylogenetically or geographically defined. \textcite{wang2015targeting}
reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
phylogenetic analysis, they found that they had ``deeply sampled'' the
local \gls{MSM} network, suggesting that many of the samples could be
epidemiologically related.

We downloaded all sequences associated with each paper from GenBank. Sequences
were multiply aligned using \software{MUSCLE} version 3.8.31, and alignments
were manually inspected with \software{Seaview} version 4.4.2. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 with the GTR model. Transmission trees
were estimated by rooting and time-scaling the phylogenies by root-to-tip
regression, using a modified version of Path-o-Gen as described
previously~\autocite{poon2015phylodynamic}.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested. For appropriate choices of the kernel meta-parameters, the classifier
had an $R^2$ above 0.8 for all combinations of tree size and epidemic size. The
$I$ parameter was also classified accurately, with an average $R^2$ of 0.93 for
500-tip trees or 0.70 for 100-tip trees. The $m$ parameter was much harder to
classify, with $R^2$ varying between 0.01 and 0.36 depending on $I$ and the
tree size. Finally, the accuracy of the classifier for $N$ varied widely, from
0.08 with 100 tips and $I = 500$, to 0.82 with 1000 tips and $I = 2000$.
Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
other parameters can be found in the supplemental materials. Based on
inspection of the cross-validation results, we chose to use the meta-parameters
$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
index or the \gls{nLTT}.

@@

@@

@@

@@

@@

\begin{figure}
  \includegraphics{leventhal2012fig1}
  \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
  \label{fig:sf1}
\end{figure}

\begin{figure}
  \includegraphics{smc_test}
  \caption{Approximation of mixture of Gaussians used by
    \textcite{del2012adaptive} and \textcite{sisson2007sequential} to test
    \gls{SMC}.}
  \label{fig:smctest}
\end{figure}

>>>>

\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}

@@

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}
\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
\newacronym{GTR}{GTR}{generalized time-reversible}

@@

\frenchspacing
\onehalfspacing
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

The structure of the contact network underlying a viral epidemic can profoundly
impact the speed and pattern of the epidemic's spread. Network structure can
influence the prevalence trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{little2014using}. True contact networks are a challenging type
of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the intra-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2011estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics}, suggesting that statistical
inference might be possible~\autocite{welch2011statistical}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[although there has been
some disagreement, see][]{jones2003assessment}. Under these models, nodes with
a high number of contacts attract new connections at an elevated rate. Networks
produced by preferential attachment have a power-law degree distribution,
meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
for some constant $\gamma$. When $2 < \gamma \leq 3$, the network is referred
to as ``scale-free''. The first contact network model incorporating
preferential attachment was introduced by \textcite{barabasi1999emergence}, and
is now referred to as the \gls{BA} model. Under this model, networks are formed
by iteratively adding nodes with $m$ new edges each. These new edges are joined
to existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
that nodes of high degree tend to attract more connections (in the original
paper, only $\alpha = 1$ was investigated).

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{brown2011transmission} fit a preferential attachment
model to a phylogenetically estimated transmission network relating 60\% of HIV
infected \gls{MSM} in the United Kingdom. The transmission network is a
subgraph of the contact network which includes only those edges which have
already led to a new infection.

Due to their complexity, it is generally difficult to explicitly calculate the
likelihood of a transmission tree under a contact network model. To do so, we
would need to integrate over all possible networks generated by that model, and
also over all possible labellings of the internal nodes of the transmission
tree. While it is not known (to us) whether such integration is tractable, a
simpler alternative is offered by likelihood-free methods, namely
\gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring}. \gls{ABC}
leverages the fact that, although calculating the likelihood may be impossible,
generating simulated datasets according to a model is often straightforward. If
our model fits the data well, the simulated data it produces should be similar
to the observed data. More formally, if $D$ is the observed data, the posterior
distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
dataset according to $\theta$, and $\varepsilon$ is a small
tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
is a kernel function, the approach is known as
kernel-\gls{ABC}~\autocite{nakagome2013kernel}.

Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
contact network parameters from an estimated transmission tree, using the tree
kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
of the \gls{BA} model on a variety of simulated and real data sets.

@@

We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics and transmission trees over static contact networks,
as has been done previously \autocite[\textit{e.g.}][]{o2010contact,
robinson2013dynamics, leventhal2012inferring, groendyke2011bayesian}. To check
that our implementation was correct, we reproduced Figure 1A of
\textcite{leventhal2012inferring} (see Figure~\ref{fig:sf1}).

We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
denotes the total number of nodes in the network, and $I$, which is the number
of infected nodes at which to stop the simulation and sample the transmission
tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
they became infected at a rate proportional to their numbers of infected
neighbours, and never recovered. For all analyses, the transmission trees'
branch lengths were scaled by dividing by their mean. We used the 
\software{igraph} library's implementation of the \gls{BA}
model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster.

@@

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. To ensure our implementation was correct, we
applied it to the same mixture of Gaussians used by
\citeauthor{del2012adaptive} to demonstrate their method \autocite[originally
used by][]{sisson2007sequential}. We were able to obtain a close approximation
to the function (see Figure~\ref{fig:smctest}), and attained the stopping
condition used by the authors in a comparable number of steps.

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1, 1.5}.
The adaptive \gls{ABC} algorithm was applied to each tree with these priors:
$m \sim$ Uniform(1, 5), $\alpha \sim$ Uniform(0, 2), and $(N, I)$ jointly
uniform on the triangular region \{$500 \leq N \leq 15000$, $500 \leq I \leq
 5000$, $I \leq N$\}. Following \textcite{del2012adaptive} and
\textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
equal to twice the empirical variance of the particles. The algorithm was run
with 1000 particles, 5 simulated datasets per particle, and the ``quality''
parameter set to 0.95. We use the same stopping criterion as
\textcite{del2012adaptive}, namely when the MCMC acceptance rate dropped below
1.5\%. Point estimates for the parameters were obtained by taking the highest
point of an estimated kernel density on the final set of particles.

@@

We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
model generates networks with a single connected component, we specifically
searched for datasets which originated from existing clusters, either
phylogenetically or geographically defined. Characteristics of the datasets we
investigated are given in Table~\ref{tab:data}.

\begin{table}
  \centering
  \begin{tabular}{ccccc}
    Reference & Sequences ($n$) & Location & Risk group & Gene \\
    \hline
    \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
    \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
    \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{1299} & \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} & \multirow{2}{*}{\textit{env}} \\
    \autocite{novitsky2014impact} \\
    \hline
  \end{tabular}
  \caption{Characteristics of published datasets investigated with kernel-ABC.
  Acronyms: MSM, men who have sex with men.}
  \label{tab:data}
\end{table}

We downloaded all sequences associated with each study from GenBank. Sequences
were multiply aligned using \software{MUSCLE} version 3.8.31
\autocite{edgar2004muscle}, and alignments were manually inspected with
\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
\gls{GTR} model. Transmission trees were estimated by rooting and time-scaling
the phylogenies by root-to-tip regression, using a modified version of
Path-o-Gen \autocite[distributed as part of \software{BEAST},
][]{drummond2007beast}, as described
previously~\autocite{poon2015phylodynamic}.

Two additional steps were undertaken for the
\textcite{novitsky2013phylogenetic, novitsky2014impact} data. First, before
performing the multiple sequence alignment, we aligned each
sequence pairwise to the HXB2 reference sequence (GenBank accession number
HIVHXB2CG) using \software{MAFFT} version 7.130b \autocite{katoh2013mafft} and
clipped out the hypervariable regions with \software{Biopython}
version 1.66+ \autocite{cock2009biopython}. Second, after multiple sequence
alignment, the sequences were trimmed to an informative region with
\software{trimAl} version 1.4.rev15 \autocite{capella2009trimal}.

Kernel-ABC was used to fit the \gls{BA} model to each dataset. To make the
analyses comparable, the same priors and configuration parameters were used for
each dataset. The priors were as follows: $\alpha \sim$ Uniform(0, 2), $m \sim$
Uniform(1, 5), and $(N, I) \sim$ Uniform(\sett{$n \leq N \leq 10000$, $n \leq I
\leq 1000$, $I \leq N$}), where $n$ is the number of sequences in the dataset.
The kernel meta-parameters and SMC configuration were the same as in the
\gls{ABC} simulations.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested. For appropriate choices of the kernel meta-parameters, the classifier
had an $R^2$ above 
    0.8 
for all combinations of tree size and epidemic size. The $I$ parameter was also
classified accurately, with an average $R^2$ of 
    0.93 
for 500-tip trees or 
    0.7 
for 100-tip trees. The $m$ parameter was much harder to classify, with $R^2$
varying between 
    0.004
and
    0.36
depending on $I$ and the tree size. Finally, the accuracy of the classifier for
$N$ varied widely, from 
    0.08
to
    0.82.
Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
other parameters can be found in the supplemental materials. Based on
inspection of the cross-validation results, we chose to use the meta-parameters
$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
index or the \gls{nLTT}.

@@

Figure~\ref{fig:abcpt} shows point estimates for the simulations with $m = 2$.
The results for the other values of $m$ are similar and can be found in the
supplemental materials.

The marginal approximated posterior distributions for one simulated dataset are
shown in Figure~\ref{fig:abcex}.

\begin{figure}
    \includegraphics{abc-point-est}
    \caption{Point estimates of \gls{BA} model parameters obtained by running
    kernel-ABC on simulated phylogenies. Dotted lines indicate the true values,
    and limits of the $y$-axes are the regions of uniform prior density.}
    \label{fig:abcpt}
\end{figure}

\begin{figure}
    \includegraphics{abc-example}
    \caption{Marginal estimated posterior distributions obtained with ABC on a
    single simulated dataset. Dotted lines and shaded polygon indicate true
    values.}
    \label{fig:abcex}
\end{figure}

\subsection*{Real data}

@@

% kernel classifiers

@@

% abc simulations

The total number of nodes $N$ was almost always significantly over-estimated.
This suggests that $N$ has very little effect on tree shape, and the estimates
were influenced primarily by the prior. Since the prior on $N$ and $I$ is
jointly uniform on a triangular region, there is more prior mass on high $N$
values. In retrospect, this is reasonable because adding more nodes to a
\gls{BA} network does not change the edge density or the network's overall
shape. This can illustrated by imagining that we grow the network after the
epidemic simulation has already been completed. It is possible (though perhaps
unlikely) that none of the new nodes attains a connection to any infected node.
Thus, running the simulation again on the new, larger network could produce the
exact same transmission tree as before. This is an extreme example, but it
serves to demonstrate the point that $N$ may be impossible to determine if the
epidemic does not come close to saturating the network.

% real data

We estimated the parameters of the \gls{BA} model for several published
datasets (see~\ref{tab:data}). \textcite{cuevas2009hiv} performed a
phylogenetic analysis of 261 patients of mixed risk groups in the Basque
Country, Spain, who were newly diagnosed between 2004 and 2007. Although their
data did not comprise a single cluster, they did find that almost half (47\%)
of the samples grouped in transmission clusters. \textcite{wang2015targeting}
reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
phylogenetic analysis, they found that they had ``deeply sampled'' the local
\gls{MSM} network, suggesting that many of the samples could be
epidemiologically related. \textcite{novitsky2013phylogenetic} investigated a
local HIV epidemic in Mochudi, Botswana. They later extended their
analysis~\autocite{novitsky2014impact} to include an estimated 70\% of HIV
infected persons in the town.

@@

\newpage

@@

\begin{figure}[ht]
  \includegraphics[scale=0.8]{leventhal2012fig1}
  \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
  \label{fig:sf1}
\end{figure}

\begin{figure}
  \includegraphics{smc_test}
  \caption{Approximation of mixture of Gaussians used by
    \textcite{del2012adaptive} and \textcite{sisson2007sequential} to test
    \gls{SMC}. Solid black line indicates true distribution. Grey shaded area
    shows SMC approximation obtained with our implementation.}
  \label{fig:smctest}
\end{figure}

### 2016-03-01 doc/thesis/results.tex
Added figures and started poster

\subsection{Separable parameters}

Our first simulation-based experiment was designed to determine which network
model parameters were potentially estimable by kernel-ABC, and to find the best
set of kernel meta-parameters to carry through to future experiments. To do
this, we simulated trees under three distinct values of each parameter of the
\gls{BA} model, and evaluated the accuracy of a kernel-\gls{SVM} classifier for
the parameter.

In Figure~\ref{fig:patrees}, we show that trees simulated under different
values of \gls{alpha} are visibly quite distinctive. In particular, higher
values of \gls{alpha} result in networks with a small number of highly
connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
likely to transmit to many other nodes. This results in a more unbalanced,
ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
all three \gls{alpha} values are well separated from each other under several
different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
1.0 using the first two principal components.

With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
classifier for \gls{alpha} was very high under a variety of prevalence and
sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
center panel). However, even for trees of size 100 sampled from an epidemic on
2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
accuracy of a Sackin's index-based classifier was also quite high, at about
0.75. The fact that Sackin's index is informative of \gls{alpha} was
unsurprising, given the clear differences in tree balance observed under
different \gls{alpha} values. The \gls{nltt}
statistic~\autocite{janzen2015approximate} did not improve the classifier, but
rather slightly reduced the $R^2$ in most cases.

@@

We also considered the possibility of inferring the number of infected nodes,
or \defn{prevalence}, under this model. All parameters except \gls{I} were
fixed at the following values: \gls{N} = 5000, \gls{alpha} = 1.0, and \gls{m} =
2. As shown in Figure~\ref{fig:prevtrees}, the prevalence had no obvious effect
on the tree shape. However, a kernel-\gls{SVM} classifier was able to
distinguish the number of infected nodes with high accuracy ($R^2 > 0.9$;
Figure~\ref{fig:prevcrossv}). Moreover, the use of the \gls{nltt} statistic
improved classification accuracy by a small amount, in contrast to the results
for \gls{alpha}. In contrast, a Sackin's index-based classifier displayed
extremely poor performance ($R^2$ < 0.1, not shown).

It is important to note that, if we cut off the epidemic simulation when 500
nodes are infected, the resulting tree will be shorter (in calendar time) than
if we continue until 2000 nodes are infected. However, this information is not
used when building the classifier, since the branch lengths in each tree are
scaled by their mean. Therefore, the high performance of the classifier is due
to structural differences captured by the tree kernel, rather than the trees
simply having different heights.

\begin{figure}[ht]
  \centering
  \label{fig:prevtrees}
  \includegraphics{kernel_nsimnode_trees}
  \caption[Similarly shaped simulated under three values of \gls{I}]{
    Epidemics were simulated on networks of size 5000 until \gls{I} = 500,
    1000, or 2000 nodes were infected. When scaled to the same height, there is
    no immediately visible distinction in shape between trees simulated under
    different \gls{I} values.
  }
\end{figure}

\begin{figure}[ht]
  \centering
  \label{fig:prevcrossv}
  \includegraphics{kernel_nsimnode_crossv}
  \caption[Cross-validation performance of kernel-\gls{SVM} classifier for
           prevalence]{
    For \gls{BA} networks of size \gls{N} = 5000, a kernel-\gls{SVM} classifier
    is very accurate at predicting \gls{I}. Epidemics were simulated until
    \gls{I} = 500, 1000, or 2000 nodes were infected, and either 100 or 500
    nodes were sampled for inclusion in a transmission tree. The parameters of
    the \gls{BA} network were \gls{alpha} = 1.0 and \gls{m} = 2. A Sackin's
    index-based classifier performed very poorly on these data ($R^2 = $ TODO
    for 100-tip trees and blah for 500-tip trees, not shown).
  }
\end{figure}

>>>>

\subsection{Kernel-SVM classifiers}

Our first simulation-based experiment was designed to determine which network
model parameters were potentially estimable by kernel-\gls{ABC}, and to find
the best set of kernel meta-parameters to carry through to future experiments.
We simulated trees under three distinct values of each parameter of the
\gls{BA} model, and calculated the cross-validation accuracy of a
kernel-\gls{SVM} classifier. Many of the figures in this section pertain to the
attachment power parameter $\alpha$ only, but equivalent figures for the other
three parameters can be found in the supplemental materials.

Trees simulated under different values of \gls{alpha} are visibly quite
distinctive (Figure~\ref{fig:patrees}). In particular, higher values of
\gls{alpha} result in networks with a small number of highly connected nodes
(see Figure~\ref{fig:pabounds}) which, once infected, are likely to transmit to
many other nodes. This results in a more unbalanced, ladder-like structure in
the phylogeny.

Kernel-\gls{PCA} projections show that all three \gls{alpha} values are well
separated from each other under several different sampling scenarios
(Figure~\ref{fig:pakpca}). For smaller trees, though, it becomes harder to
visually disinguish \gls{alpha} = 0.5 from \gls{alpha} = 1.0 using the first
two principal components.

With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
classifier for \gls{alpha} was very high under a variety of prevalence and
sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
center panel). However, even for trees of size 100 sampled from an epidemic on
2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
accuracy of a Sackin's index-based classifier was also quite high, at about
0.75. The fact that Sackin's index is informative of \gls{alpha} was
unsurprising, given the clear differences in tree balance observed under
different \gls{alpha} values.

@@

Figure~\ref{fig:rsquared} shows the cross-validation accuracy of classifiers
for all four parameters of the \gls{BA} model. The classifier for \gls{I} also
exhibited high performance, although not quite as high as that for \gls{alpha}.
The average $R^2$ values were 0.70 for 100-tip trees, or 0.93 for 500-tip
trees. On the other hand, $m$ and $N$ were much harder to classsify. For $m$,
all $R^2$ values were below 0.5, falling to virtually zero in four scenarios.
The $R^2$ values for $N$ were highly variable, ranging from 0.08 to 0.82
depending on $I$ and the number of tips in the tree. In almost all cases, both
the nLTT and Sackin's index classifiers were worse than the tree kernel, except
for classifying $N$ with $I = 2000$ and $100$ tips, where the nLTT gave a small
advantage (see the supplements). The greatest disadvantage for the Sackin's
index-based classifier was on the $I$ parameter, where it had an $R^2$ of less
than 0.1, despite the relatively high accuracy of the tree kernel (see the
supplements).

\begin{figure}[ht]
  \centering
  \includegraphics{kernel_rsquared}
  \caption[Cross-validation accuracy of kSVM classifiers for \gls{BA}
  model parameters.]{Cross-validation $R^2$ values of kernel-SVM classifier for
    \gls{BA} model parameters. Points and ranges are median and 95\% quantiles
    over 1000 2-fold cross-validations.}
  \label{fig:rsquared}
\end{figure}

% needs to go in methods

%It is important to note that, if we cut off the epidemic simulation when 500
%nodes are infected, the resulting tree will be shorter (in calendar time) than
%if we continue until 2000 nodes are infected. However, this information is not
%used when building the classifier, since the branch lengths in each tree are
%scaled by their mean. Therefore, the high performance of the classifier is due
%to structural differences captured by the tree kernel, rather than the trees
%simply having different heights.
